---
title: "Activity 4: Tidyverse"
author: "Katrina Mae A. Cabalar"
date: "01 June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load the tidyverse Package

```{r load-tidyverse}
library(tidyverse)
```

## 2. Read in the Dataset

Load the habitat types CSV file as a dataframe.

```{r load-data}
df <- read_csv("~/Desktop/habitat_types.csv")
```

## 3. Inspect the Data Structure

Use `glimpse()`, the tidyverse equivalent of `str()`, to inspect column names, data types, and sample values.

```{r glimpse}
glimpse(df)
```

## 4. Select Relevant Columns

Subset the data by keeping only Transect, Htypes, SmallTrees, and FloweringTrees.

```{r select-cols}
sel_df <- df %>%
  select(Transect, Htypes, SmallTrees, FloweringTrees)
```

## 5. Filter CVT and Sort by SmallTrees

Filter rows where the habitat type is "CVT," then sort SmallTrees in descending order to identify the transect with the most small trees.

```{r filter-cvt}
sel_df %>%
  filter(Htypes == "CVT") %>%
  arrange(desc(SmallTrees))
```

## 6. Mean Flowering Trees per Habitat Type

Calculate the average number of flowering trees for each habitat type.

```{r mean-flowering}
df %>%
  group_by(Htypes) %>%
  summarise(mean_FloweringTrees = mean(FloweringTrees))
```

## 7. Add HerbGroundDiff Column

Create a new column `HerbGroundDiff` representing the difference between HerbsCover and GroundCover. Assign the result to `sel_df_mod`, then identify transects with no difference (i.e., HerbGroundDiff == 0).

```{r herb-ground-diff}
sel_df_mod <- df %>%
  select(Transect, Htypes, SmallTrees, FloweringTrees, HerbsCover, GroundCover) %>%
  mutate(HerbGroundDiff = HerbsCover - GroundCover) %>%
  select(Transect, Htypes, SmallTrees, FloweringTrees, HerbGroundDiff)

# Transects with no difference
sel_df_mod %>%
  filter(HerbGroundDiff == 0)
```

## 8. Pivot to Wide Format

Reshape `sel_df_mod` from long to wide format so each habitat type becomes its own set of columns.

```{r pivot-wider}
sel_df_wide <- sel_df_mod %>%
  pivot_wider(
    names_from  = Htypes,
    values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff)
  )

sel_df_wide
```