---
title: "Activity 3: Summary statistics, frequency distribution, and basic graphical representation"
author: "Katrina Mae A. Cabalar"
date: "01 June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Load the Dataset

Read in the herpetozoa CSV file and preview the first few rows.

```{r load-data}
herp <- read.csv("~/Desktop/herpetozoa-033-095-s001.csv")
head(herp)
nrow(herp)
```

## 2. Clean the Dataset

Remove rows with empty Sites and rows with missing values, then confirm the new row count.

```{r clean-data}
herp <- herp[herp$Sites != "", ]
herp <- na.omit(herp)
nrow(herp)
head(herp)
str(herp)
```

## 3. Elevation Summary Statistics

Calculate the mean, standard deviation, and range of elevation across all sites.

```{r elevation-stats}
elev <- herp$Elev
mean(elev)
sd(elev)
range(elev)
```

## 4. Species Totals Using a Loop

Sum the count for each species using a for loop across species columns (columns 7 to 33).

```{r species-loop}
species_data <- herp[, 7:33]

species_totals_loop <- c()
for (i in 1:ncol(species_data)) {
  species_totals_loop[i] <- sum(species_data[, i])
}
names(species_totals_loop) <- colnames(species_data)
print(species_totals_loop)
```

## 5. Species Totals Using sapply

Repeat the same computation more concisely using `sapply()`.

```{r species-sapply}
species_totals_sapply <- sapply(species_data, sum)
species_totals_sapply
```

## 6. Most and Fewest Observed Species

Identify which species has the highest and lowest total observation count.

```{r most-fewest}
names(which.max(species_totals_sapply))
max(species_totals_sapply)

names(which.min(species_totals_sapply))
min(species_totals_sapply)

most_sp   <- names(which.max(species_totals_sapply))
fewest_sp <- names(which.min(species_totals_sapply))
```

## 7. Mean Counts by Habitat Type

Calculate the mean count of the most and fewest observed species per habitat type.

```{r habitat-means}
habitat_sp <- data.frame(
  Htypes = herp$Htypes,
  Most   = herp[, most_sp],
  Fewest = herp[, fewest_sp]
)

mean_most   <- tapply(habitat_sp$Most,   habitat_sp$Htypes, mean)
mean_fewest <- tapply(habitat_sp$Fewest, habitat_sp$Htypes, mean)

mean_most
mean_fewest

names(which.max(mean_most))
```

## 8. Scatter Plot: Elevation vs. Total Species Count

Visualize the relationship between elevation and total species count, colored by habitat type.

```{r scatter-plot}
scatter_data <- data.frame(
  Elev     = herp$Elev,
  SPPTotal = herp$SPPTotal,
  Htypes   = herp$Htypes
)

habitat_colors <- c("ASG" = "pink",
                    "ESG" = "skyblue",
                    "OGF" = "violet",
                    "CVT" = "orange")

point_colors <- habitat_colors[scatter_data$Htypes]

plot(scatter_data$Elev, scatter_data$SPPTotal,
     col  = point_colors,
     pch  = 16,
     xlab = "Elevation (m)",
     ylab = "Total Species Count (SPPTotal)",
     main = "Elevation vs. Total Species Count by Habitat Type")

legend("topright",
       legend = names(habitat_colors),
       col    = habitat_colors,
       pch    = 16,
       title  = "Habitat Type")
```

## 9. Sites with Zero Species Observations

Identify which habitat types have sites with zero observed species.

```{r zero-obs}
zero_obs <- scatter_data[scatter_data$SPPTotal == 0, "Htypes"]
cat("Habitat types with zero species observations:\n")
unique(zero_obs)
```

## 10. Bar Plot: Total Species per Habitat Type

Summarize and visualize total species counts per habitat type.

```{r barplot}
habitat_totals <- tapply(scatter_data$SPPTotal, scatter_data$Htypes, sum)

barplot(habitat_totals,
        xlab   = "Habitat Type",
        ylab   = "Total Species Count",
        main   = "Total Species Observed per Habitat Type",
        col    = c("pink", "skyblue", "violet", "orange"),
        border = "black",
        ylim   = c(0, max(habitat_totals) + 20))

text(x      = barplot(habitat_totals, plot = FALSE),
     y      = habitat_totals + 5,
     labels = habitat_totals,
     cex    = 0.9)
```

## 11. Histogram: Distribution of Total Species Counts

Examine the frequency distribution of total species counts across all sites.

```{r histogram}
hist(herp$SPPTotal,
     xlab   = "Total Species Count (SPPTotal)",
     ylab   = "Frequency (Number of Sites)",
     main   = "Histogram of Total Species Counts",
     col    = "steelblue",
     border = "white",
     breaks = 8)
```