---
title: "Activity 7: Clustering (K-means)"
author: "Katrina Mae A. Cabalar"
output: html_document
date: "08 June 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

## Part A – K-means (Single K)

### Step 1 – Read in the morphology data

```{r read-data}
morph <- read.csv("~/Desktop/morphology_data.csv")
head(morph)
```

### Step 2 – Remove the Sex column

```{r remove-sex}
morph <- morph[ , !(names(morph) %in% c("Sex"))]
head(morph)
```

### Step 3 – Confirm no NAs

```{r check-na}
sum(is.na(morph))   # should be 0
```

No missing values are present in the dataset.

### Step 4 – Separate the label column and standardize numeric variables

The `Pop` column holds the population label (character). It is set aside before 
scaling so that only the numeric morphological measurements are standardized.

```{r scale}
pop_labels <- morph$Pop                     # keep labels for later plots
morph_num  <- morph[ , !(names(morph) %in% c("Pop"))]   # numeric columns only
morph_scaled <- as.data.frame(scale(morph_num))
head(morph_scaled)
```

### Step 5 – K-means with K = 3 (25 random starts)

```{r kmeans-k3}
set.seed(42)                                 # for reproducibility
km3 <- kmeans(morph_scaled, centers = 3, nstart = 25)
```

### Step 6 – Cluster assignments and summary

```{r print-results}
# Cluster assignment for every observation
km3$cluster

# Full summary of the K-means result
print(km3)
```

### Step 7 – Scatter plot coloured by cluster (K = 3)

```{r scatter-k3, fig.width=7, fig.height=5}
library(ggplot2)

# PCA on the scaled data for a 2-D view
pca3   <- prcomp(morph_scaled)
pca3_df <- data.frame(
  PC1     = pca3$x[ , 1],
  PC2     = pca3$x[ , 2],
  Cluster = factor(km3$cluster)
)

ggplot(pca3_df, aes(x = PC1, y = PC2, colour = Cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "K-means Clustering (K = 3)",
    subtitle = "Points projected onto the first two principal components",
    x        = "PC1",
    y        = "PC2",
    colour   = "Cluster"
  ) +
  theme_bw()
```

---

## Part B – K-means (Loop)

### Step 1 – Loop over K = 2 to 10, record total within-cluster SS

```{r kmeans-loop}
set.seed(42)

wss <- numeric(9)                        # store within-SS for K = 2..10
k_values <- 2:10

for (i in seq_along(k_values)) {
  km_i      <- kmeans(morph_scaled, centers = k_values[i], nstart = 25)
  wss[i]    <- km_i$tot.withinss
}

wss_df <- data.frame(K = k_values, WSS = wss)
wss_df
```

### Step 2 – Elbow curve

```{r elbow-plot, fig.width=7, fig.height=5}
ggplot(wss_df, aes(x = K, y = WSS)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 3) +
  scale_x_continuous(breaks = k_values) +
  labs(
    title = "Elbow Curve – K-means",
    x     = "Number of Clusters (K)",
    y     = "Total Within-Cluster Sum of Squares"
  ) +
  theme_bw()
```

The elbow in the curve above occurs at **K = 3**, where the rate of decrease in
WSS slows markedly.  K = 3 is therefore selected as the optimal number of
clusters.

### Step 3 – K-means with optimal K (= 3), 25 random starts

```{r kmeans-optimal}
set.seed(42)
optimal_k  <- 3
km_opt     <- kmeans(morph_scaled, centers = optimal_k, nstart = 25)
```

### Step 4 – Cluster assignments and summary (optimal K)

```{r print-optimal}
km_opt$cluster
print(km_opt)
```

### Step 5 – Scatter plot coloured by cluster (optimal K)

```{r scatter-optimal, fig.width=7, fig.height=5}
pca_opt    <- prcomp(morph_scaled)
pca_opt_df <- data.frame(
  PC1     = pca_opt$x[ , 1],
  PC2     = pca_opt$x[ , 2],
  Cluster = factor(km_opt$cluster)
)

ggplot(pca_opt_df, aes(x = PC1, y = PC2, colour = Cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = paste0("K-means Clustering (Optimal K = ", optimal_k, ")"),
    subtitle = "Points projected onto the first two principal components",
    x        = "PC1",
    y        = "PC2",
    colour   = "Cluster"
  ) +
  theme_bw()
```
