---
title: "Activity 7: Clustering (K-means)"
author: "Katrina Mae A. Cabalar"
date: "08 June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Part A: K-means (Single K)

## Step 1: Read in the morphology data

```{r read-data}
morph <- read.csv("~/Desktop/morphology_data.csv")
head(morph)
```

## Step 2: Remove the Sex column

```{r remove-sex}
morph <- morph[ , !(names(morph) %in% c("Sex"))]
head(morph)
```

## Step 3: Check for NAs and remove if present

```{r check-na}
# Check for NAs
colSums(is.na(morph))

# Remove rows with NAs (if any)
morph <- na.omit(morph)
cat("Rows remaining after NA removal:", nrow(morph), "\n")
```

## Step 4: Standardize numeric columns using scale()

K-means is distance-based, so all numeric variables must be on the same scale.
We retain the `Pop` column for reference but scale only the numeric columns.

```{r scale}
# Separate the non-numeric Pop column
pop_col <- morph$Pop

# Select numeric columns only
morph_num <- morph[ , sapply(morph, is.numeric)]

# Standardize
morph_scaled <- as.data.frame(scale(morph_num))

# Confirm: means should be ~0 and SDs ~1
round(colMeans(morph_scaled), 4)
round(apply(morph_scaled, 2, sd), 4)
```

## Step 5: K-means with K = 3 and nstart = 25

```{r kmeans-k3}
set.seed(42)   # for reproducibility

km3 <- kmeans(morph_scaled, centers = 3, nstart = 25)
```

## Step 6: Cluster assignments and summary

```{r kmeans-output}
# Cluster assignment for each observation
cat("=== Cluster Assignments ===\n")
print(km3$cluster)

# Summary of results
cat("\n=== K-means Summary ===\n")
print(km3)
```

## Step 7: Scatter plot (SVL vs TD), coloured by cluster

```{r scatter-k3, fig.width=9, fig.height=5.5}
# Colours per cluster
cluster_colors <- c("tomato", "steelblue", "forestgreen")

# Shapes per population
pop_levels  <- c("Luzon", "Mindanao", "Palawan")
pop_shapes  <- c(16, 17, 15)   # filled circle, triangle, square
point_pch   <- pop_shapes[match(pop_col, pop_levels)]

# Expand right margin to accommodate the outside legend
par(mar = c(5, 4, 4, 10))   # bottom, left, top, right

plot(
  x    = morph_scaled$SVL,
  y    = morph_scaled$TD,
  col  = cluster_colors[km3$cluster],
  pch  = point_pch,
  xlab = "SVL (standardized)",
  ylab = "TD (standardized)",
  main = "K-means Clustering (K = 3): SVL vs TD"
)

# Allow drawing outside the plot region
par(xpd = TRUE)

# Legend 1: Clusters (colour), placed to the right of the plot
legend(
  x      = par("usr")[2] + 0.08,   # just beyond the right axis
  y      = par("usr")[4],           # top of the plot
  legend = paste("Cluster", 1:3),
  col    = cluster_colors,
  pch    = 16,
  title  = "Cluster",
  bty    = "n",
  cex    = 0.85
)

# Legend 2: Population (shape), placed below the cluster legend
legend(
  x      = par("usr")[2] + 0.08,
  y      = par("usr")[4] - 1.5,   # offset downward from cluster legend
  legend = pop_levels,
  col    = "black",
  pch    = pop_shapes,
  title  = "Population",
  bty    = "n",
  cex    = 0.85
)

# Reset xpd and margins
par(xpd = FALSE, mar = c(5, 4, 4, 2))
```

---

# Part B: K-means (Loop & Elbow Curve)

## Step 1: Loop over K = 2 to 10, record total within-cluster SS

```{r kmeans-loop}
set.seed(42)

k_values <- 2:10
wss       <- numeric(length(k_values))   # within-cluster sum of squares

for (i in seq_along(k_values)) {
  km_temp  <- kmeans(morph_scaled, centers = k_values[i], nstart = 25)
  wss[i]   <- km_temp$tot.withinss
}

# Display the WSS values
data.frame(K = k_values, WSS = round(wss, 2))
```

## Step 2: Elbow curve plot

```{r elbow-curve, fig.width=7, fig.height=5}
plot(
  x    = k_values,
  y    = wss,
  type = "b",          # both points and lines
  pch  = 19,
  col  = "steelblue",
  xlab = "Number of Clusters (K)",
  ylab = "Total Within-Cluster Sum of Squares",
  main = "Elbow Curve for K-means Clustering",
  xaxt = "n"           # suppress default x-axis to draw integer ticks
)
axis(1, at = k_values)

# Mark the elbow visually
abline(v = 3, lty = 2, col = "tomato")
text(3.15, max(wss) * 0.97, "K = 3\n(optimal)", col = "tomato", adj = 0, cex = 0.85)
```

## Step 3: K-means with optimal K (K = 3) — 25 random starts

```{r kmeans-optimal}
set.seed(42)
optimal_k <- 3

km_opt <- kmeans(morph_scaled, centers = optimal_k, nstart = 25)

# Cluster assignments
cat("=== Cluster Assignments (Optimal K =", optimal_k, ") ===\n")
print(km_opt$cluster)

# Full summary
cat("\n=== K-means Summary (Optimal K =", optimal_k, ") ===\n")
print(km_opt)
```

## Step 4: Scatter plot with optimal K

```{r scatter-optimal, fig.width=7, fig.height=5}
cluster_colors <- c("tomato", "steelblue", "forestgreen")

plot(
  x    = morph_scaled$SVL,
  y    = morph_scaled$TD,
  col  = cluster_colors[km_opt$cluster],
  pch  = 19,
  xlab = "SVL (standardized)",
  ylab = "TD (standardized)",
  main = paste0("K-means Clustering (Optimal K = ", optimal_k, "): SVL vs TD")
)

# Add cluster centroids
points(
  x   = km_opt$centers[ , "SVL"],
  y   = km_opt$centers[ , "TD"],
  col = cluster_colors,
  pch = 8,      # asterisk = centroid
  cex = 2,
  lwd = 2
)

legend(
  "topright",
  legend = c(paste("Cluster", 1:optimal_k), "Centroid"),
  col    = c(cluster_colors, "black"),
  pch    = c(rep(19, optimal_k), 8),
  bty    = "n"
)
```
