---
title: "Activity 9: Principal Component Analysis (PCA)"
author: "Katrina Mae A. Cabalar"
date: "22 June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Part A. Principal Component Analysis

## 1. Read in the Data

```{r read-data}
frogs <- read.csv("~/Desktop/Meneses_et.al_2026_Platymantis.csv", header = TRUE)
head(frogs)
```

---

## 2. Subset the Data

Retain only the `Species` column and morphological features from column 10 to the last column.

```{r subset-data}
frog_sub <- frogs[, c(1, 10:ncol(frogs))]
head(frog_sub)
```

---

## 3. Check Class and Structure

```{r check-structure}
class(frog_sub)
str(frog_sub)
```

> **How many morphological features are present?**  
> There are **`r ncol(frog_sub) - 1`** morphological features in the dataset.

---

## 4. Check for NAs

```{r check-na}
any(is.na(frog_sub))
colSums(is.na(frog_sub))
```

> The dataset has **no missing values (NAs)**. No further cleaning is required.

---

## 5. Scale the Data and Compute Pairwise Correlations

Scale only the numeric columns. The `Species` column is kept in the subset data frame but excluded from scaling and correlation.

```{r scale-correlate}
# Scale numeric columns only (columns 2 onward in the subset)
frog_scaled <- scale(frog_sub[, -1])

# Compute pairwise correlations
cor_matrix <- cor(frog_scaled)

# Preview the correlation matrix (first 6 x 6 for readability)
round(cor_matrix[1:6, 1:6], 3)
```

---

## 6. Compute Eigenvectors and Proportion of Variance

Use the correlation matrix to derive principal components manually via `eigen()`.

```{r pca-eigen}
# Compute eigenvalues and eigenvectors from the correlation matrix
eig <- eigen(cor_matrix)

# Eigenvalues
eigenvalues <- eig$values
cat("Number of principal components produced:", length(eigenvalues), "\n")

# Proportion of variance explained (in %)
prop_var <- (eigenvalues / sum(eigenvalues)) * 100
cumul_var <- cumsum(prop_var)

# Summary table
pca_summary <- data.frame(
  PC              = paste0("PC", seq_along(eigenvalues)),
  Eigenvalue      = round(eigenvalues, 4),
  Proportion_Var  = round(prop_var, 2),
  Cumulative_Var  = round(cumul_var, 2)
)
print(pca_summary)

cat("\nProportion of variance explained by PC1–PC4:\n")
for (i in 1:4) {
  cat(sprintf("  PC%d: %.2f%%\n", i, prop_var[i]))
}
```

> **How many PCs were produced?**  
> `r length(eigenvalues)` principal components were produced, one for each morphological variable.
>
> **Variance explained by PC1–PC4:**
> - PC1: `r round(prop_var[1], 2)`%  
> - PC2: `r round(prop_var[2], 2)`%  
> - PC3: `r round(prop_var[3], 2)`%  
> - PC4: `r round(prop_var[4], 2)`%  

---

## 7. Screeplot and PCA Biplot (1×2 panel)

```{r pca-biplot, fig.width=12, fig.height=6}
# Run PCA using prcomp for biplot compatibility
pca_result <- prcomp(frog_sub[, -1], scale. = TRUE)

# Define species colors and shapes
species_levels <- unique(frog_sub$Species)
sp_colors <- c("P_guiting" = "#967BB6", "P_pygmaeus" = "#9DC183")
sp_shapes <- c("P_guiting" = "20", "P_pygmaeus" = "20")

col_vec   <- sp_colors[frog_sub$Species]
shape_vec <- sp_shapes[frog_sub$Species]

# 1×2 layout
par(mfrow = c(1, 2),
    mar   = c(4, 4.5, 3, 1),
    bg    = "white",
    family = "sans")

# --- Panel 1: Screeplot ---
pct_var <- summary(pca_result)$importance[2, ] * 100

barplot(
  pct_var[1:10],
  names.arg = paste0("PC", 1:10),
  col       = ifelse(pct_var[1:10] >= mean(pct_var), "#967BB6", "#BDBDBD"),
  border    = NA,
  xlab      = "Principal Component",
  ylab      = "Variance Explained (%)",
  main      = "Screeplot",
  las       = 1,
  cex.axis  = 0.85,
  cex.names = 0.75,
  cex.lab   = 0.9,
  cex.main  = 1
)
abline(h = mean(pct_var), lty = 2, col = "#555555", lwd = 1.2)
text(x = 11.5, y = mean(pct_var) + 0.5,
     labels = "mean", col = "#555555", cex = 0.7, xpd = TRUE)

# --- Panel 2: PCA Biplot ---
scores   <- pca_result$x
loadings <- pca_result$rotation

# Scale loadings for overlay
scale_factor <- max(abs(scores[, 1:2])) / max(abs(loadings[, 1:2])) * 0.7

plot(
  scores[, 1], scores[, 2],
  col  = col_vec,
  pch  = 20,
  xlab = paste0("PC1 (", round(pct_var[1], 1), "%)"),
  ylab = paste0("PC2 (", round(pct_var[2], 1), "%)"),
  main = "PCA Biplot",
  las  = 1,
  cex  = 1.1,
  cex.axis = 0.85,
  cex.lab  = 0.9,
  cex.main = 1
)
abline(h = 0, v = 0, col = "#CCCCCC", lty = 2)

# Draw loading arrows
arrow_col <- "#3D405B80"
for (i in seq_len(nrow(loadings))) {
  arrows(
    x0 = 0, y0 = 0,
    x1 = loadings[i, 1] * scale_factor,
    y1 = loadings[i, 2] * scale_factor,
    length = 0.07, col = arrow_col, lwd = 0.9
  )
}

# Label only the top 5 loadings by PC1 magnitude to avoid clutter
top_idx <- order(abs(loadings[, 1]), decreasing = TRUE)[1:5]
text(
  loadings[top_idx, 1] * scale_factor * 1.12,
  loadings[top_idx, 2] * scale_factor * 1.12,
  labels = rownames(loadings)[top_idx],
  cex    = 0.65,
  col    = "#3D405B",
  font   = 2
)

# Legend
legend(
  "topright",
  legend = c("P. guiting", "P. pygmaeus"),
  col    = c("#967BB6", "#9DC183"),
  pch    = c(20, 20),
  bty    = "n",
  cex    = 0.85,
  pt.cex = 1.1
)
```

---

# Part B. Broken-Stick vs. PCAtest

## 1. Conceptual Comparison

> **Broken-stick method:**  
> The broken-stick model is a purely mathematical benchmark that distributes total variance across *k* components according to the expected lengths of pieces if a stick of unit length were broken at random at *k*, 1 points. A PC is considered meaningful if its eigenvalue *exceeds* the corresponding broken-stick expectation. This method requires **no data permutation or randomization** as it is a deterministic rule based on a theoretical null distribution. It is fast, reproducible, and easy to apply, but it does not provide p-values. It is a *selection criterion*, not a formal significance test.
>
> **PCAtest:**  
> PCAtest (Camargo, 2022) uses **permutation-based hypothesis testing** to assess (1) the overall significance of the entire PCA (via the Psi and Phi statistics), (2) the significance of each individual PC axis, and (3) the contribution of each variable to significant PCs. Observed eigenvalues and loading indices are compared against null distributions built by randomly shuffling the data many times. This produces actual **p-values** and bootstrap-derived confidence intervals. It is computationally more intensive but statistically rigorous.
>
> **Key difference:** The broken-stick method tells you which PCs *retain more variance than expected by chance* under a simple geometric model; PCAtest tells you which PCs are *statistically significant* based on the actual multivariate correlation structure of the data, with formal p-values.

---

## 2. Broken-Stick Analysis (Barplot)

```{r broken-stick, fig.width=8, fig.height=5}
# Broken-stick expected values for k components
broken_stick_expected <- function(k) {
  sapply(1:k, function(i) {
    (1 / k) * sum(1 / (i:k))
  }) * 100  # as percentage of total variance
}

k          <- length(pca_result$sdev)
bs_exp     <- broken_stick_expected(k)
obs_var    <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100

# Build a comparison data frame (show first 10 PCs)
n_show <- 10
bs_df <- data.frame(
  PC       = rep(paste0("PC", 1:n_show), 2),
  Variance = c(obs_var[1:n_show], bs_exp[1:n_show]),
  Type     = rep(c("Observed", "Broken-stick"), each = n_show)
)

# Plot side-by-side bars
bar_positions <- barplot(
  matrix(c(obs_var[1:n_show], bs_exp[1:n_show]), nrow = 2, byrow = TRUE),
  beside      = TRUE,
  names.arg   = paste0("PC", 1:n_show),
  col         = c("#9DC183", "#BDBDBD"),
  border      = NA,
  xlab        = "Principal Component",
  ylab        = "Variance Explained (%)",
  main        = "Broken-Stick Analysis: Observed vs. Expected Variance",
  las         = 1,
  cex.axis    = 0.85,
  cex.names   = 0.75,
  cex.lab     = 0.9,
  cex.main    = 0.95
)

legend(
  "topright",
  legend = c("Observed", "Broken-stick expected"),
  fill   = c("#9DC183", "#BDBDBD"),
  bty    = "n",
  border = NA,
  cex    = 0.85
)

# Indicate which PCs exceed broken-stick threshold
exceed <- which(obs_var[1:n_show] > bs_exp[1:n_show])
cat("PCs where observed variance exceeds the broken-stick expectation:", 
    paste(paste0("PC", exceed), collapse = ", "), "\n")
```

---

## 3. PCAtest – Permutation-Based Significance Test

```{r pcatest-run, fig.width=7, fig.height=6}
library(PCAtest)

# Run PCAtest on the scaled numeric morphological data
# Arguments: data matrix, nboot, nperm, alpha, varcorr, counter, plot
pca_test_result <- PCAtest(
  x        = frog_sub[, -1],   # numeric morphological variables only
  nboot    = 1000,              # bootstrap replicates
  nperm    = 1000,              # permutation replicates
  alpha    = 0.05,              # significance level
  varcorr  = FALSE,             # use raw variance (not variance corrected)
  counter  = FALSE,             # suppress iteration counter
  plot     = TRUE               # produce PCAtest diagnostic plot
)

# Print summary
print(pca_test_result)
```

> **Interpreting PCAtest output:**  
> - The **Psi** and **Phi** statistics assess whether there is more correlation structure in the data than expected by chance. A p-value < 0.05 means the overall PCA is significant.  
> - For each PC, the test compares the observed eigenvalue against the maximum eigenvalue from random permutations. PCs with p < 0.05 are considered significant.  
> - For significant PCs, the loading index identifies which variables contribute meaningfully.

---