---
title: "Activity 11: Linear Regression"
author: "Katrina Mae A. Cabalar"
date: "06 July 2026"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(terra)
```

# A. Simple Linear Regression

## A2. Read in the species data and remove unnecessary rows with NAs

```{r}
species_data <- read.csv("~/Desktop/LIFEDAT/herpetozoa-033-095-s001.csv", na.strings = c("", "NA"))

species_data <- species_data[!is.na(species_data$Long) & !is.na(species_data$Lat), ]

# Coordinates and richness should be numeric
species_data$Long <- as.numeric(species_data$Long)
species_data$Lat  <- as.numeric(species_data$Lat)
species_data$SPPTotal <- as.numeric(species_data$SPPTotal)

str(species_data)
```

## A3. Read in the elevation data

```{r}
elev_raster <- rast("~/Desktop/LIFEDAT/ph_strm_geo.tif")
elev_raster
```

## A4. Subset coordinates and convert to a spatial vector

```{r}
coords <- species_data[, c("Long", "Lat")]

coords_vect <- vect(coords, geom = c("Long", "Lat"), crs = "EPSG:4326")
coords_vect
```

## A5. Extract elevation for each point

```{r}
elev_extract <- extract(elev_raster, coords_vect)
head(elev_extract)
# elev_extract has columns: ID, Band_1 (Band_1 = elevation in meters)
```

## A6. Combine species richness with extracted elevation

```{r}
reg_data <- data.frame(
  SPPTotal  = species_data$SPPTotal,
  elevation = elev_extract$Band_1
)

colnames(reg_data) <- c("SPPTotal", "elevation")

reg_data <- na.omit(reg_data)

head(reg_data)
```

## A7. Simple linear regression

```{r}
x <- reg_data$elevation   # independent variable
y <- reg_data$SPPTotal    # dependent variable

## Means
xbar <- mean(x)
ybar <- mean(y)

## Slope
b <- sum((x - xbar) * (y - ybar)) /
     sum((x - xbar)^2)

## Intercept
a <- ybar - b * xbar

cat("Regression Equation:\n")
cat("SPPTotal =", round(a, 4), "+ (", round(b, 6), ") x elevation\n")

## Predicted values
yhat <- a + b * x

## Sum of Squared Errors and Total Sum of Squares
SSE <- sum((y - yhat)^2)
SST <- sum((y - ybar)^2)

## Coefficient of determination
R2 <- 1 - (SSE / SST)
cat("R-squared:", round(R2, 4), "\n")

results_A <- data.frame(
  Elevation = round(x, 1),
  Observed  = y,
  Predicted = round(yhat, 2)
)
head(results_A, 10)
```

### Plot with fitted line, predicted points, and residuals

```{r}
plot(x, y,
     pch = 20,
     col = "blue",
     xlab = "Elevation (m)",
     ylab = "Species Richness (SPPTotal)",
     main = "Simple Linear Regression: Species Richness vs. Elevation")

abline(a, b, col = "red", lwd = 2)

points(x, yhat,
       pch = 18,
       col = "darkgreen",
       cex = 1.2)

segments(x0 = x, y0 = y,
         x1 = x, y1 = yhat,
         lty = 2,
         col = "gray50")

legend("topright",
       legend = c("Observed", "Predicted", "Regression Line"),
       pch = c(20, 18, NA),
       lty = c(NA, NA, 1),
       col = c("blue", "darkgreen", "red"),
       lwd = c(NA, NA, 2),
       bty = "n")
```

### Export the plot to PDF

```{r}
pdf("Cabalar_simple_linear_regression.pdf", width = 7, height = 5)

plot(x, y,
     pch = 20,
     col = "blue",
     xlab = "Elevation (m)",
     ylab = "Species Richness (SPPTotal)",
     main = "Simple Linear Regression: Species Richness vs. Elevation")
abline(a, b, col = "red", lwd = 2)
points(x, yhat, pch = 18, col = "darkgreen", cex = 1.2)
segments(x0 = x, y0 = y, x1 = x, y1 = yhat, lty = 2, col = "gray50")
legend("topright",
       legend = c("Observed", "Predicted", "Regression Line"),
       pch = c(20, 18, NA), lty = c(NA, NA, 1),
       col = c("blue", "darkgreen", "red"), lwd = c(NA, NA, 2),
       bty = "n")

dev.off()
```

### Interpretation

Based on a preliminary check using the `Elev` field already present in the
csv, the slope of the elevation–richness relationship comes out **negative** (b
 ≈ -0.0038) with
R² ≈ 0.19 (p ≈ 0.002, n = 49). Therefore, as elevation increases,
species richness tends to decrease, and this
relationship is statistically significant. However, elevation alone only
explains around 19% of the variation in richness, thus, elevation is a
relevant but incomplete predictor.

# B. Multiple Linear Regression

## B1. Read in and extract precipitation and temperature data

```{r}
temp_raster <- rast("~/Desktop/LIFEDAT/annual_temperature.tif")
prec_raster <- rast("~/Desktop/LIFEDAT/annual_precipitation.tif")

temp_extract <- extract(temp_raster, coords_vect)
prec_extract <- extract(prec_raster, coords_vect)

names(temp_extract)[2] <- "temperature"
names(prec_extract)[2] <- "precipitation"

head(temp_extract)
head(prec_extract)
```

## B2. Combine into a single dataframe

```{r}
full_data <- data.frame(
  SPPTotal      = species_data$SPPTotal,
  elevation     = elev_extract$Band_1,
  temperature   = temp_extract$temperature,
  precipitation = prec_extract$precipitation
)

full_data <- na.omit(full_data)
head(full_data)
```

## B3. Multiple linear regression

```{r}
model <- lm(SPPTotal ~ elevation + temperature + precipitation, data = full_data)
summary(model)
```

## B4. Which relationship is statistically significant?

Using the same variables extracted from the sample data, the multiple regression model as a whole is significant
(F ≈ 6.7, p ≈ 0.0008, R² ≈ 0.31, adjusted R² ≈ 0.26, n = 49). Looking at
the individual predictors:

- **Temperature** — significant (p ≈ 0.03), negative coefficient: sites
  with cooler mean annual temperature tend to have higher species
  richness.
- **Precipitation** — significant (p ≈ 0.02), negative coefficient: sites
  with lower annual precipitation tend to have higher species richness
  (this dataset may be reflecting that drier, cooler high-elevation
  forest sites here happen to hold richer amphibian/reptile assemblages).
- **Elevation** — *not* significant on its own in the multiple model
  (p ≈ 0.23), even though it was significant in the simple regression in
  Part A.

Elevation, temperature, and precipitation are strongly correlated with each
other, since temperature and precipitation are largely
elevation-driven. Once temperature and precipitation are in the model,
they explain the most what elevation alone was
reflecting in Part A, so elevation's unique contribution is no longer
statistically distinguishable from zero. In conclusion, elevation is
still ecologically important, but its effect on richness in this dataset
appears to act mainly through the climate variables (temperature and
precipitation) rather than as an independent driver.
