---
title: "Activity 8: Hierarchical Clustering"
author: "Katrina Mae A. Cabalar"
date: "21 August 2026"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(cluster)
library(factoextra)
library(dendextend)
```

## Read and prep data

```{r data}
morpho <- read.csv("~/Desktop/LIFEDAT/morphology_data.csv", stringsAsFactors = FALSE)
str(morpho)

pop_labels <- morpho$Pop

morpho_num <- morpho %>%
  select(-Pop, -Sex)
str(morpho_num)

sum(is.na(morpho_num))
complete_rows <- complete.cases(morpho_num)
morpho_num  <- morpho_num[complete_rows, ]
pop_labels  <- pop_labels[complete_rows]
sum(is.na(morpho_num))
```

## Distance matrix

```{r distance}
morpho_scaled <- scale(morpho_num)
euc_dist <- dist(morpho_scaled, method = "euclidean")
```

## Hierarchical clustering: Ward's and Single Linkage

```{r hclust}
hc_ward   <- hclust(euc_dist, method = "ward.D2")
hc_single <- hclust(euc_dist, method = "single")
```

## Silhouette score across K = 2 to 10

```{r silhouette-loop}
k_values <- 2:10

sil_results <- data.frame(
  K = rep(k_values, times = 2),
  Method = rep(c("Ward", "Single"), each = length(k_values)),
  Avg_Silhouette = NA_real_
)

row_i <- 1
for (method_hc in list(Ward = hc_ward, Single = hc_single)) {
  for (k in k_values) {
    clusters_k <- cutree(method_hc, k = k)
    sil <- silhouette(clusters_k, euc_dist)
    sil_results$Avg_Silhouette[row_i] <- mean(sil[, "sil_width"])
    row_i <- row_i + 1
  }
}

sil_results
```

```{r silhouette-plot}
ggplot(sil_results, aes(x = K, y = Avg_Silhouette, color = Method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = k_values) +
  labs(
    title = "Average Silhouette Score by K",
    subtitle = "Ward's vs. Single Linkage",
    x = "Number of Clusters (K)",
    y = "Average Silhouette Width"
  ) +
  theme_minimal()
```

```{r optimal-k}
best_k_ward   <- sil_results %>% filter(Method == "Ward")   %>% slice_max(Avg_Silhouette, n = 1) %>% pull(K)
best_k_single <- sil_results %>% filter(Method == "Single") %>% slice_max(Avg_Silhouette, n = 1) %>% pull(K)

best_k_ward
best_k_single
```

## Dendrograms with optimal K highlighted

```{r dendro-ward, fig.height=6, fig.width=10}
fviz_dend(
  hc_ward,
  k = best_k_ward,
  cex = 0.5,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  main = paste0("Ward's Method Dendrogram (Optimal K = ", best_k_ward, ")")
)
```

```{r dendro-single, fig.height=6, fig.width=10}
fviz_dend(
  hc_single,
  k = best_k_single,
  cex = 0.5,
  rect = TRUE,
  rect_fill = TRUE,
  rect_border = "jco",
  k_colors = "jco",
  main = paste0("Single Linkage Dendrogram (Optimal K = ", best_k_single, ")")
)
```

## K-means vs. Hierarchical

```{r kmeans-rerun}
best_k_kmeans <- 3

set.seed(42)
km_result <- kmeans(morpho_scaled, centers = best_k_kmeans, nstart = 25)
km_result$size
```

```{r kmeans-comparison}
hc_ward_clusters   <- cutree(hc_ward,   k = best_k_kmeans)
hc_single_clusters <- cutree(hc_single, k = best_k_kmeans)

table(KMeans = km_result$cluster, Ward = hc_ward_clusters)
table(KMeans = km_result$cluster, Single = hc_single_clusters)
```

```{r agreement-check}
if (!require(mclust, quietly = TRUE)) install.packages("mclust")
library(mclust)

adjustedRandIndex(km_result$cluster, hc_ward_clusters)
adjustedRandIndex(km_result$cluster, hc_single_clusters)
```

```{r compare-to-pop}
table(Pop = pop_labels, KMeans = km_result$cluster)
table(Pop = pop_labels, Ward   = hc_ward_clusters)
table(Pop = pop_labels, Single = hc_single_clusters)
```

## Are the clusters consistent?

**What to expect going in.** Activity 7's K-means (K = 3, seed 42) split
the 202 individuals into groups of 73, 34, and 95, explaining about 43.7%
of the total variance (between_SS / total_SS). `Pop` itself splits as
Luzon = 93, Mindanao = 72, Palawan = 37 -- close to, but not identical to,
the K-means group sizes, which is already a hint that the morphological
clusters don't map onto population one-to-one.

**Ward's linkage vs. K-means.** Because Ward's method minimizes
within-cluster variance at each merge (much like K-means minimizes
within-cluster sum of squares), the two usually agree reasonably well when
the data have compact, roughly-spherical groups.

**Single linkage vs. K-means.** Single linkage merges clusters based on the
*closest pair* of points between them, which makes it prone to "chaining"
-- a few intermediate individuals can string two otherwise distinct groups
into one long cluster, or one dominant cluster absorbs almost everything
while a couple of outliers get isolated as their own tiny "cluster." If the
Single-linkage cross-tab shows one huge cluster and one or two
singleton/near-singleton clusters, that's chaining in action -- and it will
show up as a low ARI against K-means even though the two methods aren't
necessarily "wrong," just sensitive to different things.

**If the clusters are not consistent, what's driving it:**

- **Different objective functions.** K-means and Ward's both try to
  minimize within-group variance, but K-means re-optimizes centroids
  iteratively and can escape local groupings hierarchical merging can't
  undo; Ward's makes irreversible merge decisions bottom-up. Single linkage
  optimizes for nothing but nearest-neighbor distance, so it isn't really
  compatible with the other two.
- **Randomness in initialization.** K-means depends on the starting
  centroids; `nstart = 25` mitigates this but doesn't eliminate all
  sensitivity, whereas hierarchical clustering is fully deterministic given
  a distance matrix and linkage rule.
- **Shape assumptions.** K-means assumes clusters are roughly spherical and
  similar in size/variance; if the true morphological groups are elongated,
  nested, or very unequal in spread, K-means can badly split or merge them
  in ways hierarchical clustering (especially Ward's) may not.

**Why this matters biologically.** If Ward's and K-means agree reasonably
well with each other but neither agrees particularly well with `Pop`, that
suggests the *morphological* variation in this dataset doesn't line up
cleanly with population/locality -- i.e., morphology may be shaped more by
sex, individual variation, or measurement noise than by population-level
divergence. Conversely, if Single linkage disagrees sharply with both
K-means and Ward's, that disagreement is more likely a methodological
artifact (chaining) than a real alternative biological signal, and it would
be a mistake to read biological meaning into it without first checking for
that artifact. In general: when two conceptually similar methods (K-means,
Ward's) agree but a third, less-comparable method (Single) disagrees, trust
the agreement between the first two more than the outlier -- and always
cross-check any clustering result against a known grouping variable like
`Pop` before treating it as evidence of a "real" biological pattern.
